```{r}
library(censusapi)
library(tidyverse)
Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")
```

```{r}
load("C:/Users/derek/Google Drive/City Systems/Stockton Green Economy/acs5_vars_2017.Rdata")
```

EMPLOYMENT STATUS
```{r}
acs5subjectemp <-
  getCensus(
    name = "acs/acs5/subject",
    vintage = 2018,
    vars = c("S2301_C01_001E","S2301_C02_001E","S2301_C03_001E","S2301_C04_001E"),
    region = "county:077",
    regionin = "state:06"
  ) %>%
  mutate(
    Population16andOlder = S2301_C01_001E,
    PercLaborForceParticipation = S2301_C02_001E,
    PercEmployedResidents = S2301_C03_001E,
    LaborForceParticipation = PercLaborForceParticipation/100*Population16andOlder,
    EmployedResidents = PercEmployedResidents/100*Population16andOlder,
    UnemploymentRate = S2301_C04_001E
  ) %>%
  dplyr::select(
    Population16andOlder,
    PercLaborForceParticipation,
    PercEmployedResidents,
    LaborForceParticipation,
    EmployedResidents,
    UnemploymentRate
  )
```

MEANS OF TRANSPORTATION TO WORK BY CLASS OF WORKER
```{r}
acs5emp1 <-
  getCensus(
    name = "acs/acs5",
    vintage = 2017,
    vars = c("B08128_001E","B08128_002E","B08128_003E","B08128_004E"),
    region = "county:077",
    regionin = "state:06"
  ) %>% 
  select_if(!names(.) %in% c("GEO_ID","state","county","NAME")) %>%
  dplyr::select(-c(contains("EA"),contains("MA"),contains("M"))) %>%
  gather(
    key = "variable",
    value = "estimate"
  ) %>%
  mutate(
    label = acs_vars$label[match(variable,acs_vars$name)]
  ) %>% 
  select(-variable)
```

FULL-TIME, YEAR-ROUND WORK STATUS IN THE PAST 12 MONTHS BY AGE FOR THE POPULATION 16 YEARS AND OVER.
```{r}
acs5emp2 <-
  getCensus(
    name = "acs/acs5",
    vintage = 2017,
    vars = "group(B23027)",
    region = "county:077",
    regionin = "state:06"
  ) %>% 
  select_if(!names(.) %in% c("GEO_ID","state","county","NAME")) %>%
  dplyr::select(-c(contains("EA"),contains("MA"),contains("M"))) %>%
  gather(
    key = "variable",
    value = "estimate"
  ) %>%
  mutate(
    label = acs_vars$label[match(variable,acs_vars$name)]
  ) %>% 
  select(-variable) %>% 
  separate(
    label,
    into = c(NA,NA,NA,"work","work2"),
    sep = "!!"
  )

tot_work <-
  acs5emp2 %>% 
  filter(!is.na(work2)) %>% 
  dplyr::select(estimate) %>% 
  sum()

tot_fulltime <-
  acs5emp2 %>% 
  filter(work2=="Worked full-time year-round") %>% 
  dplyr::select(estimate) %>% 
  sum()
```

MARITAL STATUS BY SEX BY LABOR FORCE PARTICIPATION
Best match so far -- only off by a few hundred (within rounding error). Seems to be counting armed forces, which S2301 might include. 
```{r}
acs5emp3 <-
  getCensus(
    name = "acs/acs5",
    vintage = 2017,
    vars = "group(B12006)",
    region = "county:077",
    regionin = "state:06"
  ) %>% 
  select_if(!names(.) %in% c("GEO_ID","state","county","NAME")) %>%
  dplyr::select(-c(contains("EA"),contains("MA"),contains("M"))) %>%
  gather(
    key = "variable",
    value = "estimate"
  ) %>%
  mutate(
    label = acs_vars$label[match(variable,acs_vars$name)]
  ) %>% 
  select(-variable) %>% 
  separate(
    label,
    into = c(NA,NA,"marital","sex","participate","work"),
    sep = "!!"
  )

tot_labor <-
  acs5emp3 %>% 
  filter(participate == "In labor force") %>% 
  filter(is.na(work)) %>% 
  select(estimate) %>% 
  sum()

tot_employed_or_military <-
  acs5emp3 %>% 
  filter(work == "Employed or in Armed Forces") %>% 
  select(estimate) %>% 
  sum()
```

EMPLOYMENT STATUS FOR THE POPULATION 16 YEARS AND OVER
Seems to be best match. S2301 "Labor Force Participation" appears to include armed forces. "Employment" does not include armed forces, which is what we would want for this analysis anyway.
```{r}
acs5emp4 <-
  getCensus(
    name = "acs/acs5",
    vintage = 2018,
    vars = "group(B23025)",
    region = "county:077",
    regionin = "state:06"
  ) %>% 
  select_if(!names(.) %in% c("GEO_ID","state","county","NAME")) %>%
  dplyr::select(-c(contains("EA"),contains("MA"),contains("M"))) %>%
  gather(
    key = "variable",
    value = "estimate"
  ) %>%
  mutate(
    label = acs_vars$label[match(variable,acs_vars$name)]
  ) %>% 
  select(-variable) %>% 
  separate(
    label,
    into = c(NA,NA,"participate","type","work"),
    sep = "!!"
  )

tot_labor <-
  acs5emp4 %>% 
  filter(participate == "In labor force") %>% 
  filter(is.na(type)) %>% 
  select(estimate) %>% 
  sum()

tot_employed <-
  acs5emp4 %>% 
  filter(work == "Employed") %>% 
  select(estimate) %>% 
  sum()
```

```{r}
acs5emp4 <-
  getCensus(
    name = "acs/acs5",
    vintage = 2018,
    vars = "group(B23025)",
    region = "block group:*",
    regionin = "state:06+county:077"
  ) %>% 
  mutate(
      bg = paste0(state,county,tract,block_group)
  ) %>% 
  select_if(!names(.) %in% c("GEO_ID","state","county","tract","block_group","NAME")) %>%
  dplyr::select(-c(contains("EA"),contains("MA"),contains("M"))) %>%
  gather(
    key = "variable",
    value = "estimate",
    -bg
  ) %>%
  mutate(
    label = acs_vars$label[match(variable,acs_vars$name)]
  ) %>% 
  select(-variable) %>% 
  separate(
    label,
    into = c(NA,NA,"participate","type","work"),
    sep = "!!"
  )

tot_labor <-
  acs5emp4 %>% 
  filter(participate == "In labor force") %>% 
  filter(is.na(type)) %>% 
  select(estimate) %>% 
  sum()

tot_employed <-
  acs5emp4 %>% 
  filter(work == "Employed") %>% 
  select(estimate) %>% 
  sum()

employed_bgs <-
  acs5emp4 %>% 
  filter(work == "Employed")

sjc_bgs <- block_groups("CA") %>% 
  filter(COUNTYFP == "077")

sjc_bgs %>% 
  left_join(
    employed_bgs,
    by = c("GEOID"="bg")
  ) %>% 
  mapview(zcol="estimate")
```